---
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r echo = FALSE}
options(width = 100)
suppressPackageStartupMessages(suppressMessages(library(dplyr)))
```

# ps

> List, Query, Manipulate System Processes

[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis build status](https://travis-ci.org/r-lib/ps.svg?branch=master)](https://travis-ci.org/r-lib/ps)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/r-lib/ps?branch=master&svg=true)](https://ci.appveyor.com/project/gaborcsardi/ps)
[![CRAN status](https://www.r-pkg.org/badges/version/ps)](https://cran.r-project.org/package=ps)
[![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/ps)](https://www.r-pkg.org/pkg/ps)
[![Coverage status](https://codecov.io/gh/r-lib/ps/branch/master/graph/badge.svg)](https://codecov.io/github/r-lib/ps?branch=master)

## Introduction

ps implement an API to query and manipulate system processes. Most of its
code is based on the [psutil](https://github.com/giampaolo/psutil) Python
package.

## Installation

Once released, you can install the released version of ps from
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("ps")
```

## Supported platforms

ps currently supports Windows (from Vista), macOS and Linux systems.
On unsupported platforms the package can be installed and loaded, but
all of its functions fail with an error of class `"not_implemented"`.

## Listing all processes

`ps_pids()` returns all process ids on the system. This can be useful to
iterate over all processes.

```{r}
library(ps)
ps_pids()[1:20]
```

`ps()` returns a data frame (tibble if you have the tibble package
available), with data about each process. It contains a handle to each
process, in the `ps_handle` column, you can use these to perform more
queries on the processes.

```{r}
ps()
```

## Process API

This is a short summary of the API. Please see the documentation of the
various methods for details, in particular regarding handles to finished
processes and pid reuse. See also "Finished and zombie processes"
and "pid reuse" below.

`ps_handle(pid)` creates a process handle for the supplied process id.
If `pid` is omitted, a handle to the calling process is returned:

```{r}
p <- ps_handle()
p
```

### Query functions

`ps_pid(p)` returns the pid of the process.

```{r}
ps_pid(p)
```

`ps_create_time()` returns the creation time of the process (according to
the OS).

```{r}
ps_create_time(p)
```

The process id and the creation time uniquely identify a process in a
system. ps uses them to make sure that it reports information about, and
manipulates the correct process.

`ps_is_running(p)` returns whether `p` is still running. It handles pid
reuse safely.

```{r}
ps_is_running(p)
```

`ps_ppid(p)` returns the pid of the parent of `p`.

```{r}
ps_ppid(p)
```

`ps_parent(p)` returns a process handle to the parent process of `p`.

```{r}
ps_parent(p)
```

`ps_name(p)` returns the name of the program `p` is running.

```{r}
ps_name(p)
```

`ps_exe(p)` returns the full path to the executable the `p` is running.

```{r}
ps_exe(p)
```

`ps_cmdline(p)` returns the command line (executable and arguments) of `p`.

```{r}
ps_cmdline(p)
```

`ps_status(p)` returns the status of the process. Possible values are OS
dependent, but typically there is `"running"` and `"stopped"`.

```{r}
ps_status(p)
```

`ps_username(p)` returns the name of the user the process belongs to.

```{r}
ps_username(p)
```

`ps_uids(p)` and  `ps_gids(p)` return the real, effective and saved user
ids of the process. They are only implemented on POSIX systems.

```{r}
if (ps_os_type()[["POSIX"]]) ps_uids(p)
if (ps_os_type()[["POSIX"]]) ps_gids(p)
```

`ps_cwd(p)` returns the current working directory of the process.

```{r}
ps_cwd(p)
```

`ps_terminal(p)` returns the name of the terminal of the process, if any.
For processes without a terminal, and on Windows it returns `NA_character_`.

```{r}
ps_terminal(p)
```

`ps_environ(p)` returns the environment variables of the process.
`ps_environ_raw(p)` does the same, in a different form. Typically they
reflect the environment variables at the start of the process.

```{r}
ps_environ(p)[c("TERM", "USER", "SHELL", "R_HOME")]
```

`ps_num_threads(p)` returns the current number of threads of the process.

```{r}
ps_num_threads(p)
```

`ps_cpu_times(p)` returns the CPU times of the process, similarly to
`proc.time()`.

```{r}
ps_cpu_times(p)
```

`ps_memory_info(p)` returns memory usage information. See the manual for
details.

```{r}
ps_memory_info(p)
```

`ps_children(p)` lists all child processes (potentially recuirsively) of
the current process.

```{r}
ps_children(ps_parent(p))
```

### Process manipulation

`ps_suspend(p)` suspends (stops) the process. On POSIX it sends a SIGSTOP
signal. On Windows it stops all threads.

`ps_resume(p)` resumes the process. On POSIX it sends a SIGCONT signal. On
Windows it resumes all stopped threads.

`ps_send_signal(p)` sends a signal to the process. It is implemented on
POSIX systems only. It makes an effort to work around pid reuse.

`ps_terminate(p)` send SIGTERM to the process. On POSIX systems only.

`ps_kill(p)` terminates the process. Sends `SIGKILL` on POSIX systems,
uses `TerminateProcess()` on Windows. It make an effort to work around
pid reuse.

## Finished and zombie processes

ps handles finished and Zombie processes as much as possible.

The essential `ps_pid()`, `ps_create_time()`, `ps_is_running()` functions
and the `format()` and `print()` methods work for all processes, including
finished and zombie processes. Other functions fail with an error of class
`"no_such_process"` for finished processes.

The `ps_ppid()`, `ps_parent()`, `ps_children()`, `ps_name()`,
`ps_status()`, `ps_username()`, `ps_uids()`, `ps_gids()`, `ps_terminal()`,
`ps_children()` and the signal sending functions work properly for
zombie processes. Other functions fail with `"zombie_process"` error.

## Pid reuse

ps functions handle pid reuse as well as technically possible.

The query functions never return information about the wrong process, even
if the process has finished and its process id was re-assigned.

On Windows, the process manipulation functions never manipulate the wrong
process.

On POSIX systems, this is technically impossible, it is not possible to
send a signal to a process without creating a race condition. In ps the
time window of the race condition is very small, a few microseconds, and
the process would need to finish, _and_ the OS would need to reuse its pid
within this time window to create problems. This is very unlikely to
happen.

## Recipes

In the spirit of [psutil recipes](http://psutil.readthedocs.io/en/latest/#recipes).

### Find process by name

Using `ps()` and dplyr:

```{r}
library(dplyr)
find_procs_by_name <- function(name) {
  ps() %>%
    filter(name == !!name)  %>%
    pull(ps_handle)
}

find_procs_by_name("R")
```

Without creating the full table of processes:

```{r}
find_procs_by_name <- function(name) {
  procs <- lapply(ps_pids(), function(p) {
    tryCatch({
      h <- ps_handle(p)
      if (ps_name(h) == name) h else NULL },
      no_such_process = function(e) NULL,
      access_denied = function(e) NULL
    )
  })
  procs[!vapply(procs, is.null, logical(1))]
  }

find_procs_by_name("R")
```

### Wait for a process to finish

On POSIX, there is no good way to wait for non-child processes to finish,
so we need to write a sleep-wait loop to do it. (On Windows, and BSD
systems, including macOS, there are better solutions.)

```{r}
as_secs <- function(x) as.numeric(x, units = "secs")

wait_for_process <- function(proc, timeout = Inf, sleep = 0.1) {
  sleep <- as_secs(sleep)
  deadline <- Sys.time() + timeout
  while (ps_is_running(proc) && (timeout == Inf || Sys.time() < deadline)) {
    to <- min(as_secs(deadline - Sys.time()), sleep)
    Sys.sleep(to)
  }
  ! ps_is_running(proc)
}

px <- processx::process$new("sleep", "2")
p <- ps_handle(px$get_pid())
wait_for_process(p, 1)
wait_for_process(p)
```

### Wait for several processes to finish

This is similar, but we need to wait on all processes in a loop.

```{r}
wait_for_processes <- function(procs, timeout = Inf) {
  gone <- list()
  alive <- procs
  deadline <- Sys.time() + timeout

  check_gone <- function(proc, timeout) {
    proc_gone <- wait_for_process(proc, timeout = timeout)
    if (proc_gone) {
      gone <<- c(gone, list(proc))
      alive <<- setdiff(alive, list(proc))
    }
  }

  while (length(alive)) {
    if (timeout <= 0) break
    for (proc in alive) {
      max_timeout <- 1 / length(alive)
      if (timeout != Inf) {
        timeout <- min(as_secs(deadline - Sys.time()), max_timeout)
        if (timeout <= 0) break
        check_gone(proc, timeout)
      } else {
        check_gone(proc, max_timeout)
      }
    }
  }
  list(gone = gone, alive = alive)
}

px1 <- processx::process$new("sleep", "10")
px2 <- processx::process$new("sleep", "10")
px3 <- processx::process$new("sleep", "1")
px4 <- processx::process$new("sleep", "1")

p1 <- ps_handle(px1$get_pid())
p2 <- ps_handle(px2$get_pid())
p3 <- ps_handle(px3$get_pid())
p4 <- ps_handle(px4$get_pid())

wait_for_processes(list(p1, p2, p3, p4), timeout = 2)
```

### Kill process tree

This sends a signal, so it'll only work on Unix. Use `ps_kill()`
instead of `ps_send_signal()` on Windows.

```{r}
kill_proc_tree <- function(pid, sig = signals()$SIGTERM,
                           include_parent = TRUE) {
  if (pid == Sys.getpid() && include_parent) stop("I refuse to kill myself")
  parent <- ps_handle(pid)
  children <- ps_children(parent, recursive = TRUE)
  if (include_parent) children <- c(children, parent)
  for (p in children) ps_send_signal(p, sig)
  wait_for_processes(children, timeout = 0.1)
}

p1 <- processx::process$new("sleep", "10")
p2 <- processx::process$new("sleep", "10")
p3 <- processx::process$new("sleep", "10")
kill_proc_tree(Sys.getpid(), include_parent = FALSE)
```

### Terminate children

Note, that some R IDEs, including RStudio, run a multithreaded R process,
and other threads may start processes as well. `reap_children()` will clean
up all these as well, potentially causing the IDE to misbehave or crash.

```{r}
reap_children<-function(timeout = 3) {
  procs<-ps_children(ps_handle())

  ## SIGTERM
  lapply(procs, ps_terminate)

  ga<-wait_for_processes(procs, timeout = timeout)

  ## SIGKILL to the survivers
  if (length(ga$alive)) lapply(ga$alive, ps_kill)

  ga2<-wait_for_processes(ga$alive, timeout = timeout)

  ## Some might still survive
  list(gone = c(ga$gone, ga2$gone), alive = ga2$alive)
}

pxs <- replicate(3, processx::process$new("sleep", "3"))
reap_children()
```

### Filtering and sorting processes

Process name ending with "sh":

```{r}
ps() %>%
  filter(grepl("sh$", name))
```

Processes owned by user:

```{r}
ps() %>%
  filter(username == Sys.info()[["user"]]) %>%
  select(pid, name)
```

Processes consuming more than 100MB of memory:

```{r}
ps() %>%
  filter(rss > 100 * 1024 * 1024)
```

Top 3 memory consuming processes:

```{r}
ps() %>%
  top_n(3, rss) %>%
  arrange(desc(rss))
```

Top 3 processes which consumed the most CPU time:

```{r}
ps() %>%
  mutate(cpu_time = user + system) %>%
  top_n(3, cpu_time) %>%
  arrange(desc(cpu_time))
```

## Contributions

Please note that this project is released with a
[Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).
By participating in this project you agree to abide by its terms.

## License

BSD © RStudio
