---
output: github_document
---

# ps

> List system processes

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/r-lib/ps.svg?branch=master)](https://travis-ci.org/r-lib/ps)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/r-lib/ps?branch=master&svg=true)](https://ci.appveyor.com/project/r-lib/ps)
[![CRAN status](https://www.r-pkg.org/badges/version/ps)](https://cran.r-project.org/package=ps)
[![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/ps)](https://www.r-pkg.org/pkg/ps)
[![Coverage status](https://codecov.io/gh/r-lib/ps/branch/master/graph/badge.svg)](https://codecov.io/github/r-lib/ps?branch=master)

## Installation

Once released, you can install the released version of ps from
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("ps")
```

## Listing all processes

`ps_pids()` returns all process ids on the system. This can be useful to
iterate over all processes.

```{r}
library(ps)
ps_pids()[1:20]
```

`ps()` is similar, but it returns a data frame (tibble if you have the
tibble package available), with data about each process:

```{r}
ps()
```

## Process API

TODO

## Low level process API

TODO

## Process cleanup

TODO

## processx and ps

TODO

## Contributions

Please note that this project is released with a
[Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md).
By participating in this project you agree to abide by its terms.

## License

BSD © RStudio
